---
title: "Courts in Liberal Democracy"
subtitle: "Week 2"
author: "Dr Michal Ovádek"
search: false
format:
  revealjs:
    logo: img/ucl-logo.png
    footer: "POLS0113: Judicial Politics"
    slide-number: c/t
    show-slide-number: all
    center: true
    margin-left: "0.1"
    margin-right: "0.1"
bibliography: ../references.bib
---

```{r globals, warning=FALSE, message=FALSE, error=FALSE}
source("utils/globals.R")
```

## Why should unelected judges have the last word?

-   this week:

    -   why anyone would empower a court

    -   the political theory that frames the module: rule of law, separation of powers, judicial review

    -   the concept doing most of the work, **judicial independence**

-   what does it take for a court to constrain a government that could simply ignore it?

# Why courts get power

## Why would anyone empower a court?

-   the creation and maintenance of powerful courts is often explained by **credible commitments**

-   by empowering courts outside their control, governments signal willingness to be bound by constitutional rules (e.g. property rights) [@north1989constitutions]

    -   internationally, governments signal commitment to treaties, which new democracies use to **lock in** their preferences [@moravcsik2000origins]

## Why would anyone empower a court? {.smaller}

| theory | mechanism | who empowers courts | evidence |
|------|------|------|------|
| credible commitment | binding oneself reassures investors | rulers seeking investment | @north1989constitutions |
| insurance | courts protect tomorrow's losers | incumbents expecting defeat | @ginsburg2003judicial |
| hegemonic preservation | threatened elites lock in their values | dominant but declining elites | @hirschl2004political |
| lock-in | tying the hands of future governments | new democracies | @moravcsik2000origins |

::: notes
Insurance and hegemonic preservation predict opposite empowerers: those about to lose power vs those still dominant but worried about the future. Ginsburg and Versteeg (2014) test several of these against each other.
:::

## Why would anyone empower a court? {.smaller}

| theory | mechanism | who empowers courts | evidence |
|------|------|------|------|
| diffusion | copying neighbours and models | constitution drafters | @ginsburg2014why |
| functionalist | complex governance needs adjudication | nobody in particular | |
| rights revolution | litigants push from below | civil society | @epp1998rights |
| policy entrenchment | parties insulate policies from elections | governing parties | @gillman2002how |

# The rule of law

## What is the rule of law? {.smaller}

-   **thin** (formal) conceptions: law must be general, public, prospective, clear, consistent, possible to comply with, stable and actually applied (Fuller's eight desiderata, 1964)

    -   says nothing about the content of the law; an unjust law can satisfy all eight

-   **thick** (substantive) conceptions add rights, democracy or social justice

    -   the more you pack in, the less the concept can explain

-   courts feature in both: as the guarantee of consistent application (thin) and as the guardians of rights (thick)

-   @waldron2021rule takes a procedural view: what matters is the hearing and the argument

## Rule of law in comparative perspective

```{r vdemrule}
vdemdata::plot_indicator("v2x_rule") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_continuous() +
  labs(y = "Rule of law",
       title = "Rule of law",
       subtitle = "V-Dem index 'rule of law'")
```

## Rule of law in comparative perspective

```{r vdemrulectrs}
vdemdata::plot_indicator("v2x_rule",
                         min_year = 1990,
                         countries = c("Hungary", "Poland", "United Kingdom", "Venezuela")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_continuous() +
  labs(y = "Rule of law",
       title = "Rule of law",
       subtitle = "V-Dem index 'rule of law'") +
  guides(linetype = "none")
```

# Separation of powers and judicial review

## Separation of powers

-   even in systems where judicial independence is strong, there are many formal points of contact between the judiciary, the legislature and the executive

    -   appointment of judges

    -   competence disputes

    -   judicial review of administrative acts

    -   judicial review of statutory law

## Designs of constitutional review {.smaller}

-   **who reviews**: one constitutional court (centralized) or every court (diffuse)

    -   centralized: Austria, Germany, Italy, most of post-1989 Europe; diffuse: US, Argentina, Japan, the Nordic countries

-   **when**: before promulgation (*a priori*, France until 2008) or after (*a posteriori*, almost everywhere else)

-   **how**: in the abstract, detached from any case (referral by politicians), or concretely, in a dispute between parties

-   **who can trigger it**: politicians, courts, ombudsmen, individuals (constitutional complaint in Germany, Spain or Hungary; the *tutela* in Colombia)

-   **how strong**: strike down (strong form) or declare incompatible and leave parliament to act (weak form)

    -   weak form: UK Human Rights Act 1998, Canada, New Zealand

-   each dimension changes which cases arrive and who can be blamed for the result

## *Ex ante* v *ex post* review

-   one type of indirect effect of judicialization is legislator moral hazard [@fox2011judicial; @ward2019judicial]

    -   knowledge that courts will correct ill-advised policies makes legislators behave more recklessly

-   this type of moral hazard is more common with *ex ante* judicial review [@gabel2024complaw]

::: notes
French Constitutional Council is the archetype of the ex ante review effect on legislative recklessness

legislators promising unconstitutional policies can make it harder for voters to make good decisions
:::

## The counter-majoritarian difficulty {.smaller}

-   Bickel (1962): an unelected court striking down the acts of elected majorities

| position | core claim | who |
|------|------|------|
| rights as trumps | some questions are beyond the reach of majorities | Dworkin |
| process theory | review polices the democratic process | Ely (1980) |
| the uneasy case | review reduces rights violations, even if undemocratic | Fallon (2008) |
| the case against | legislatures should settle disagreements about rights | @waldron2021rule |

-   all four are about what courts *should* do; most of this module is about what they *do*

## Is the court counter-majoritarian in practice? {.smaller}

-   Dahl's answer in 1957: the US Supreme Court rarely blocks a lawmaking majority for long, because appointments bring it back into line [@dahl1957decisionmaking]

-   the difficulty may be smaller in practice than in theory

# Judicial power and independence

## Who cares about courts?

-   it is commonly assumed that court rulings must be complied with and followed by affected actors in virtually all constitutional systems

-   but what are the sources of this *judicial power*? why and when does anyone listen to courts?

## Concept of judicial power

-   following @staton2011judicial, we can usefully identify two pillars of judicial power:

    -   **autonomy** – or independence in a strict sense – or the extent to which judges' decisions reflect their true preferences as opposed to someone else's

    -   **effectiveness** – or compliance – or the extent to which courts can compel actors to comply with adverse decisions

## Concept of judicial power

|                   |                        |                       |
|-------------------|------------------------|-----------------------|
|                   | **High Effectiveness** | **Low Effectiveness** |
| **High Autonomy** | Powerful Court         | Ignorable Court       |
| **Low Autonomy**  | Puppet Court           | Joke Court            |

: The two dimensions of judicial power

## Judicial power in comparative perspective

```{r vdem}
vdemdata::plot_indicator("v2x_jucon") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_continuous() +
  scale_y_continuous(limits = c(0.25,0.75)) +
  labs(y = "Judicial Power",
       title = "Judicial Power",
       subtitle = "V-Dem index 'judicial constraints on the executive'")
```

## Judicial power in comparative perspective

```{r vdemctrs}
vdemdata::plot_indicator("v2x_jucon",
                         min_year = 1990,
                         countries = c("Hungary", "United States of America", "Mexico")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_continuous() +
  labs(y = "Judicial Power",
       title = "Judicial Power",
       subtitle = "V-Dem index 'judicial constraints on the executive'") +
  guides(linetype = "none")
```

## What makes judges autonomous?

-   **judicial independence** is the single most important attribute associated with modern judiciaries

-   yet there is disagreement on the precise content of this concept

-   independence to do what? independence from what or whom?

## Judicial independence

> the degree to which judges actually decide cases in accordance with their own determinations of the evidence, the law and justice, free from the coercion, blandishments, interference, or threats from governmental authorities or private citizens. [@rosenn1987protection, p. 8]

## What is(n't) judicial independence

-   can a judge be independent when the judiciary as a whole isn't?

-   does "dependence" on another judge's or court's decision undermine judicial independence?

## What is(n't) judicial independence

-   what about anticipatory judicial behaviour?

    -   a court may self-censor if it knows the government is likely to override its decision

-   free from "governmental authorities **or** private citizens"

-   rulings based on "own determinations of ... the law" vs "the law"

## De jure vs de facto independence

-   many scholars distinguish between "paper" or institutional guarantees of independence and what happens in practice [@melton2014does; @hayo2007explaining]

-   the most common *de jure* guarantees are:

    -   commitment to judicial independence, life tenure, depoliticized selection procedure, constrained removal procedure, salary non-regression

## De jure vs de facto independence

-   there is mixed evidence on the importance of *de jure* guarantees of independence on *de facto* levels

    -   some find *de jure* independence to matter little in the aggregate [@melton2014does], others find it to be the most important predictor of *de facto* independence [@hayo2007explaining]

-   specific guarantees nonetheless seem to change behaviour when they change, tenure and reappointment rules in particular

## De jure vs de facto independence

-   there is doubt whether judicial autonomy can be measured in isolation due to anticipation effects

    -   judges' decisions are affected by the likelihood of compliance [@linzer2015global]

## Measuring judicial independence {.smaller}

-   three families of measures:

    -   **de jure**: code the constitution and statutes (tenure, appointment, removal, salary) – cheap and comparable, but paper

    -   **de facto**: ask experts (V-Dem, CIRI) or read behaviour (rulings against the government) – closer to reality, but subjective or contaminated

    -   **latent**: combine many noisy indicators into one estimate with a measurement model [@linzer2015global]

-   the available measures disagree with each other more than their users assume [@riosfigueroa2014evaluation]

-   behaviour is a treacherous indicator: a court that never rules against the government may be captured, or may never be asked to

    -   and a government that expects to lose may not legislate at all

## How can courts increase their power?

-   new courts consolidate power by not antagonizing the government for an initial period

-   expand jurisdiction through broad interpretation of access rules and legal domains

-   a common strategy is to first create the power-expanding principle without applying it in the initial case to mitigate costs of compliance (*Marbury*)

## Why would anyone in power build a court that can rule against them?

## References
