---
title: "The Impact of Judicial Decisions"
subtitle: "Week 9"
author: "Dr Michal Ovádek"
search: false
format:
  revealjs:
    logo: img/ucl-logo.png
    footer: "POLS0113: Judicial Politics"
    slide-number: c/t
    show-slide-number: all
    center: true
    margin-left: "0.1"
    margin-right: "0.1"
bibliography: ../references.bib
---

```{r globals, warning=FALSE, message=FALSE, error=FALSE}
source("utils/globals.R")
```

## Do court rulings change anything?

-   courts command neither the sword nor the purse, yet their rulings are treated as changing the world

-   this week: how decisions get executed, how much of their effect survives the politics, and what they do to behaviour, the economy and public opinion

-   how would we know that a ruling caused anything?

# Two streams

## How do decisions turn into outcomes? {.smaller}

-   two **streams** through which judicial decisions get executed

-   the **conveyor belt**: the relevant bureaucracy (civil service, police, prisons, registries) implements the decision mechanically

    -   execution depends on the **quality of implementing institutions**: capacity, professionalism, corruption

    -   the same quality determines how much compliance there is

    -   most decisions and most of the impact courts have, but rarely studied

-   the **contested** stream: some decisions become contested or politicized

    -   execution still runs through the implementing institutions

    -   but impact now also depends on **other political actors** (executive, legislature, parties, other courts) and on **public support**

    -   compliance becomes a choice between implementation, evasion, override and backlash

::: notes
What moves a decision from one stream to the other? Salience, distributive stakes, who loses, media attention. The shadow effect sits upstream of both: some decisions never need executing because policy was adapted in anticipation. Support deterring non-compliance is the public-support arrow in the lower stream.
:::

## Two streams of judicial impact

```{mermaid}
%%| fig-width: 10
flowchart LR
  D[judicial decision] --> Q{contested or<br/>politicized?}
  Q -->|no| I1[implementing institutions<br/>civil service, police, agencies]
  I1 -->|institutional quality| O1[execution and compliance]
  Q -->|yes| I2[implementing institutions]
  Q -->|yes| P[political actors<br/>executive, legislature, parties]
  Q -->|yes| S[public support]
  S -.-> P
  I2 --> O2[compliance, evasion,<br/>override or backlash]
  P --> O2
  S --> O2
```

# Compliance

## Compliance: concept and measurement {.smaller}

-   **compliance**: the addressee does what the ruling requires; **implementation**: the ruling is put into practice by whoever has to; **impact**: anything changes as a result, intended or not

-   two types of effects of courts in judicialized political systems: **direct** effects of litigation and rulings, **indirect** effects of anticipation and threat

-   measuring compliance is harder than it looks: rulings differ in what they demand, addressees comply partially or late, and "full compliance" is often a judgment call [@kapiszewski2013compliance]

-   and compliance is endogenous: courts write the rulings they expect to be obeyed, so observed compliance rates overstate judicial power

## (Non-)Compliance

-   due to courts' enforcement-dependency, non-compliance is an attractive strategy for other actors (especially other branches of power)

    -   but non-compliance can itself generate more litigation

-   in most cases it is less costly for the government/legislator to evade compliance implicitly

    -   e.g. creating a parliamentary working group but never delivering [@vanberg2001legislativejudicial]

## How courts secure compliance {.smaller}

-   **vagueness**: judges write vaguer opinions where they fear defiance, trading clarity for the appearance of compliance [@staton2008value]

-   **publicity**: public hearings and press work raise the cost of ignoring a ruling [@krehbiel2016politics]

-   **monitoring**: the Costa Rican Supreme Court built a compliance monitoring system with follow-up orders [@gauri2015costa]

-   **domestic politics**, for international courts: new democracies comply faster with the ECtHR [@grewal2015are]; compliance depends on domestic institutions and coalitions [@hillebrecht2014power]; media attention to a ruling predicts whether it is complied with [@reis2025media]; leaders defy the Inter-American Court when their publics oppose the remedy [@parente2025domestic]

-   in the conveyor-belt stream compliance depends on the bureaucracy, not on these tools

## Judgments waiting to be implemented {.smaller}

-   at the end of 2024, 3,916 judgments of the ECtHR were pending full implementation, 1,149 of them "leading" cases that require a change in law or practice (Council of Europe annual report, 2024)

-   894 cases were closed during the year: the stock is not growing, but it is not shrinking fast either

-   cases against Russia, expelled from the Council of Europe in 2022, make up about 44% of the stock, with more than three billion euros of awards unpaid

-   the states with the most pending leading cases combine a heavy caseload with slow legislation (Russia, Ukraine, Turkey); Romania is the counter-example of building domestic capacity to implement

-   implementation of international judgments runs through domestic parliaments, ministries and courts

# The hollow hope

## The hollow hope debate {.smaller}

-   **the constrained court**: courts lack the tools to produce social change; *Brown* did little for desegregation until Congress and the executive acted a decade later [@rosenberg2023hollow]

-   **the dynamic court**: litigation works indirectly, by legitimizing claims, mobilizing movements and raising rights consciousness [@mccann1992reform]

-   **the backlash thesis**: *Brown* mattered, but by radicalizing Southern politics and provoking the violence that made civil rights legislation possible [@klarman1994how]

-   three claims about the same ruling; the disagreement is partly about which outcome counts as impact and over what horizon

## Four kinds of impact {.smaller}

|  | material | symbolic |
|------|------|------|
| **direct** | the addressees do what the ruling orders: policy adopted, prisoners released, payments made | the parties and the public come to frame the issue as a matter of rights |
| **indirect** | actors not addressed change behaviour: agencies, firms, other courts; mobilization is triggered | the wider public debate and the media reframe the problem |

-   typology from the study of Colombia's T-025 ruling on displaced people [@rodriguezgaravito2015radical]

-   most impact studies measure the top-left cell; the other three are where Rosenberg and McCann disagree

## Process tracing {.smaller}

-   some effects of courts leave no ruling behind

    -   e.g. policy changed in anticipation of review (the **shadow effect**)

-   **process tracing** follows one case step by step through documents, interviews and timing

    -   a Norwegian welfare reform preempted review by a little-known international court [@pavone2022shadow]

    -   political entrepreneurs captured Guatemala's courts through the nomination process [@moralesforte2025entrepreneurial]

-   does each step leave the predicted evidence? does a rival explanation fit too?

-   one case reveals a mechanism; @martinsen2015judicial pairs case studies with a dataset

## Structural rulings {.smaller}

-   some rulings order a whole policy rather than a remedy for one claimant

    -   Colombia's T-025 on displaced people (2004), Brazil's declaration of an "unconstitutional state of affairs" in its prisons (2015), South Africa's order to roll out antiretroviral treatment (2002), India's right-to-food orders (from 2001)

-   the court keeps the case open and supervises: follow-up hearings, reporting duties, deadlines

-   implementation runs through ministries and budgets, so these rulings sit in the contested stream from the start

-   the evidence on their effects is mixed: large symbolic and indirect effects, slow and partial material ones [@rodriguezgaravito2015radical]

## The limits of judicialization {.smaller}

-   judicial empowerment is usually pursued with the hope of increasing government accountability and individual right protection

-   @botero2022limits argue that in Latin America, judicialization mostly failed to deliver on these promises due to

    -   weak state capacity limiting the desired positive impact of courts

    -   weak social embeddedness of (liberal) legal structures

    -   the capture of courts by special interests (corruption, nepotism)

::: notes
judicialization in Latin America as a product of the 1980s and 1990s liberal (either domestic or foreign) visions, out of step with

in Lava Jato Brazilian prosecutors and judges were coming up with creative legal solutions; this opens them up even further to criticism
:::

# What rulings do

## What rulings do {.smaller}

| ruling or decision | outcome | effect |
|----------------------------|------------------------|------------------------------------------------|
| *Dobbs* (2022) | births | up about 2.3% in states that banned abortion [@dench2024effects] |
| *Dobbs* | trust in the Supreme Court | a lasting fall |
| *Shelby County* (2013) | Black turnout | fell relative to white turnout in formerly covered areas [@billings2024disparate] |
| *Citizens United* (2010) | elections | Republican gains |
| *Obergefell* (2015) | social norms | shifted; personal attitudes did not |
| CJEU *Junqueras* (2019) | euroscepticism in Spain | rose |

## Outside spending before and after *Citizens United*

```{r citizensunited}
read_csv("data/outside_spending.csv", show_col_types = FALSE) |>
  ggplot(aes(x = cycle, y = spending_bn, fill = type)) +
  geom_col() +
  geom_vline(xintercept = 2010, linetype = "dashed") +
  annotate("text", x = 2010.2, y = 4.2, label = "Citizens United\n(January 2010)", hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = seq(2006, 2024, 2)) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  labs(x = NULL, y = "Billion USD (nominal)",
       title = "Outside spending in US federal elections",
       subtitle = "Spending by groups other than candidates and party committees, by election cycle",
       caption = "Source: OpenSecrets, as reported by The Conversation (2006-2016) and in press coverage of the 2018-2024 cycles; 2024 as of 16 November 2024")
```

## *Citizens United* (2010) {.smaller}

-   January 2010: the Supreme Court held that corporations and unions may spend unlimited sums on independent political advertising; two months later *SpeechNow* opened the way to super PACs

-   outside spending by non-party groups went from a few hundred million dollars per cycle to billions; part of the rise is money that would have flowed through other channels anyway, part of it is new

-   the ruling struck down bans in about half the states, which supplies a comparison: Republican candidates gained in the states that had bans relative to those that did not [@klumpp2016business]

-   caveats for the chart: nominal dollars, definitions of outside spending differ between sources, and the 2024 figure covers only part of the cycle

## Judges' decisions and defendants' lives {.smaller}

-   the most literal impact of a judicial decision is on the person in the dock

-   as-if random assignment of judges turns a judge's leniency into an instrument for what happens next

    -   Norway: incarceration reduced reoffending and raised employment, through training and treatment in prison [@bhuller2020incarceration]

    -   US: pretrial detention raised the probability of conviction and lowered employment years later [@dobbie2018effects]

-   the same design gives opposite results; what a sentence does depends on what the sentence is

## Courts and the economy {.smaller}

-   in the classic narrative, independent courts constrain the executive's discretion, including to engage in arbitrary coercive acts, which benefits economic development [@north1989constitutions]

    -   across a sample of 104 countries, de facto (but not de jure) judicial independence is associated with higher growth rates [@voigt2015economic]

    -   externally funded judicial reforms in African countries improved perception of the rule of law among relatively powerless groups [@chemin2021can]

-   court quality has firm-level effects: comprehensive judicial reforms raise the productivity of firms in contract-intensive sectors [@chemin2020judicial]; less congested courts amplified the effect of Brazil's bankruptcy reform on credit and investment [@ponticelli2016court]

-   these are effects of judicial institutions rather than of particular rulings

# Rulings and public opinion

## Four public reactions to a ruling {.smaller}

-   **acceptance and legitimation**: a trusted court moves opinion towards the policy it upheld [@sternberg2022legitimacyconferring], helped by legalistic media framing [@linos2016supreme]; courts may be no better at this than other institutions [@cohen2024judicialization]

-   **backlash**: the ruling moves opinion against it, as in Catalonia [@casas2024judicial]; *Brown* [@klarman1994how]; yet opinion backlash is rarer than assumed [@bishin2016opinion]

-   **nothing**: *Bush v Gore* left the Court's legitimacy intact [@gibson2024losing]; public mood responds to the Court's direction thermostatically at best [@ura2014backlash]

-   **spillover**: the ruling changes what people think of the court or the polity rather than of the policy, as after *Dobbs* [@levendusky2024has; @gibson2025do], *Junqueras* [@turnbulldugarte2022can] and unpopular UK Supreme Court rulings [@ovadek2026contagion]


## Support for SCOTUS after *Dobbs* {.smaller}

-   seemingly even controversial decisions such as *Bush v. Gore* did not significantly reduce diffuse support for SCOTUS [@gibson2024losing]

-   ... that is until *Dobbs* [@levendusky2024has]

-   the fall in legitimacy lingered at least six months, though it weakened over time [@gibson2025do]

-   the incongruity for positivity theory: legitimacy was supposed to be resilient to unpopular rulings

## Support for SCOTUS post-*Dobbs*

![Effect of the Dobbs ruling on trust in SCOTUS [@levendusky2024has]](img/levendusky-dobbs.jpg)

## Rulings and attitudes {.smaller}

-   legalization of same-sex marriage reduced anti-gay bias in the states that legalized it [@kreitzer2014does; @ofosu2019samesex]

-   *Obergefell* moved what people believed others thought, not what they themselves thought: rulings can shift **norms** without shifting attitudes [@tankard2017effect]

-   the opinion backlash that many predicted after gay-rights advances did not materialize [@bishin2016opinion]

-   the durable effects of rulings on attitudes run through norms and behaviour more than through persuasion

## Do rulings change society, or follow changes already under way?

## References
